---
title: "Group Project"
author: "Laura Angley, Laura Brubaker-Wittman, Christian Gagnon, Mel Zarate"
date: "11/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Christian is the coolest member of this group!

Packages we need: 
```{r}
library(curl)
library(MASS)
library(candisc)
library(ggplot2)
```


```{r}
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN597_Fall19/KamilarAndCooperData.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = TRUE)
head(d)
summary(d)
```

LDA
```{r}
kamcoop.lda <- lda(formula = Genus ~ Body_mass_male_mean + Body_mass_female_mean + Mass_Dimorphism, data = d)
kamcoop.lda
```
The coefficient here is super small, but I'm not sure what this means.

Predict function generate value from selected model function. The length of the value predicted will be correspond with the length of the processed data.
```{r}
lda.predict <- predict(kamcoop.lda)
lda.predict
#ldahist(lda.predict) #there should be an argument g = and I'm not sure what it should be set to (should be an object). It should make histograms of the lda results
```
Making a scatter plot- this is a direct copy and paste from the wine data example that I will change later: 
```{r}
plot(lda.predict$x[,1], lda.predict$x[,2])
```

